### 2020-07-12 document/manuscript-out.tex
Travis build: 131

Access to literature is of paramount importance in the process of
scientific inquiry. The free flow of ideas and information are crucial
to addressing economic, environmental and social issues facing
developing countries especially in Africa. However, contrary to
concerted research efforts to foster development, an assessment of
Africa's research output between 1999 - 2008 indicated that Africa
lagged behind in it's research output when compared to the rest of the
world \cite{adams2010;raju2012}. Poor accessibility of scientific
publications has been suggested as one of the reasons for the low
research output from Africa \cite{tise2011;raju2012}.

>>>>

Access to literature is of paramount importance in the process of
scientific inquiry. The free flow of ideas and information are crucial
to addressing economic, environmental and social issues facing
developing countries especially in Africa. However, contrary to
concerted research efforts to foster development, an assessment of
Africa's research output between 1999 - 2008 indicated that Africa
lagged behind in it's research output when compared to the rest of the
world \cite{adams2010;raju2012}. Poor accessibility of scientific
publications has been suggested as one of the reasons for the low
research output from Africa \cite{tise2011;raju2012}. Accessibility of
scientific literature by African researchers is mostly hindered by the
lack of access to subscription-based journal articles due to their
unaffordable cost. However, the main barrier to equitable access and
sharing of knowledge in Africa is the cost publishing, which is
unaffordable for most researchers in Low- and Middle-income countries.

The concept of openness in scientific research has long been proposed as
a means toward equitable sharing of knowledge
\cite{burton_sharing_2011;nikos_building_2019}. Consequently, most
journals have shifted from subscription-based access to research
articles and embraced open access publishing. In this case, however, the
authors have to pay article publication fees which often run into
thousands of dollars - another barrier to publishing especially from
authors in Low- and Middle-income countries. In most Kenyan
institutions, for example, students have to publish at least one paper
at MSc level. With limited funds and the need to publish in a reputable
journal, most students usually don't know which journals to publish in
due to the high cost of article processing charges (APCs).

In this study, we observed that Kenyan authors are increasingly adopting
open access publishing, a trend that has been observed in Low Income
countries \cite{iyandemye_low_2019}. To help students and early career
scientists adopt this open access publishing trend, we looked into
publishers that publish open access and created a resource that acts as
a guideline for them to publish open access, and at low cost. We found
that most Open Access publishers offer waivers or subsidies to authors
from developing countries. The publishers had different criteria for
offering waivers. Some publishers, such as Nature Publishing Group,
offered a full waiver to authors in low-income countries and 50\%
discount for authors in low-middle-income countries. Some of the
publishers' waiver policies are linked to the corresponding author,
source of funds for research or the financial need of the author. For
example, PLOS offers waivers of about 50\% or more if the research was
carried out in low or LMIC countries therefore linking their waiver
policy to source of research funds. Publishers such as Oxford University
Press and eLIFE offer waivers to authors in low- and middle-income
countries who experience financial hardships or have insufficient funds.
Waivers are also available on a case-by-case basis for authors in low-
and middle-income countries by publishers such as MDPI.

As a low-middle-income country, Kenyan authors are eligible for waivers
and subsidies. Although we found no special waiver to students
identified so far, students may apply for waivers as authors in
financial need in developing countries whom publishers offer waivers on
a case-by-case basis. Additionally, students may also place their
manuscript on preprint servers, the green route of publishing, where
they can receive feedback on their work before publication in an open
access journal. Both the use of preprint servers and open access
publishing by taking advantage of the waivers offered by publishers will
increase the accessibility of research literature in Africa.
